---
title: "Introduction to Tidyverse Module"
author: "Quinn Thomas"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This module represent the "Whole Game" for data science in R. In this module you will be importing, tidying, transforming, visualizing, and communicating environmental data.  If you have used Tidyverse previously, then this module is a review.  If you have not used Tidyverse previously then you will use the links to the "R for Data Science" book to learn the skills that you need to complete the module.  

Be sure to load the following packages

```{r}
#install packages("tidyverse")
library(tidyverse)
library(lubridate)
```

The Reservoir column is different reservoirs in the Roanoke, VA area that a set of water quality variables have been measured.

- Description of data set: <https://portal.edirepository.org/nis/metadataviewer?packageid=edi.200.12>
- Data: <https://pasta.lternet.edu/package/data/eml/edi/200/12/0a62d1946e8d9a511bc1404e69e59b8c>

**Warming-up**

Read the data file into R. This file is directly streamed in R memory from a url, rather than downloaded and saved locally.

```{r}
#df <- read_csv("https://pasta.lternet.edu/package/data/eml/edi/200/12/0a62d1946e8d9a511bc1404e69e59b8c")

file_name <- "https://pasta.lternet.edu/package/data/eml/edi/200/12/0a62d1946e8d9a511bc1404e69e59b8c"

download.file(url = file_name, destfile = "data_raw/assignment2_water_quality_raw.csv", method = "curl")


df <- read_csv(“data_raw/assignment2_water_quality_raw.csv”)

```

# Excerise 1

The first exercise is focused on data visualization so you will start with a filtered dataset named `df2`.  This subsetted dataset is for a particular location (`Site == 50`) for a particular Reservoir (`Reservoir == "FCR"`)

```{r}
df2 <- df |> 
  filter(Reservoir == "FCR" & Site == 50 & Date == as_datetime("2021-05-28 10:48:31"))

#how to find number of rows and columns
nrow(df)
nrow(df2)
ncol(df)
ncol(df2)
```

## Question 1:

- Describe the data set that was imported above (`df`) to the best of your ability given the documentation provided: what are the columns and units? Where do the numbers come from?

#Q1 Answer: Within df, there are 996363 observations. There are 25 columns: Reservoir (name of reservoir), Site (site number of each data point), Date (ymd & time), Depth (meters), Temperature (celcius), dissolved oxygen (mgl), dissolved oxygen (pSat), conductivity (uscm), specific conductance (uScm), chlorophyll-a (ugL), water turbidity, water pH, reduction potential (mV), Photosynthetically active radiation (umolm2s), descent rate (m/s). The numbers are data collected from techologies of specific points at a site in a reservoir of a certain characteristic (such as depth, temperature, surface chlorophyll-a. 

- What value is used for missing values?  What is the interpretation of missing values?

#Q2 Answer: The value "NA" is used for missing values. "NA" stands for nonapplicable, so anytime a function is run on any NA values, R cannot compute and it will output another NA.

## Question 2

[Plot](https://r4ds.hadley.nz/data-visualize.html#two-numerical-variables) and *describe* the relationship between temperature and depth.  What might explain the pattern? Hint: use the point geom

```{r}
#Gives a quick view of the data table
head(df2)

ggplot(df2, aes(x = Depth_m, y = Temp_C)) + 
  geom_point() +
  labs(x = "Depth (m)", y = "Water Temperature (C)") +
  theme_classic()

#Answer: The water temperature closest to the surface is typically greater than the water temperature at higher depths. This might be explained by the water's exposure to the sun. The water at the surface is in more direct sunlight therefore being exposed to more heat from the sun, whereas the water at higher depths are gradually being exposed to less and less sunlight, which explains the downward curve of the graph as depth increases. 

```

# Exercise 2

Exercise 2 is focused on cleaning up the full dataset.  Initially, you will be using the full dataset (`df`).  You will be using the `filter`, `select`, `rename`, and `mutate` functions.

## Question 3

We are only interested in the `Chla_ugL` at the surface of each reservoir. [Filter](https://r4ds.hadley.nz/data-transform.html#filter) the dataset (`df`) to only include measurements less than 1 m deep and [select](https://r4ds.hadley.nz/data-transform.html#sec-select) only the reservoir, site, date, depth, and chlorophyll-a columns.   Remember the filtering is done on rows and selecting is done on columns.

## Question 4

The column name `Date` is not an accurate name for the column because it include the time of day as well.  [Rename](https://r4ds.hadley.nz/data-transform.html#rename) the column to `datetime` and save the updated data frame. 

```{r}
#3
df3 <- df |> filter(Depth_m < 1) |> select(Reservoir:Depth_m, Chla_ugL)

#4
df3 <- df3 |> rename("datetime" = `Date`)

```


## Question 5

In the analysis below, you will need a column that is the year.  Apply the function [`year()`](https://r4ds.hadley.nz/datetimes.html#date-time-components) from the lubridate package to the `datetime` column and save the updated data frame. Hint: you will use the `year` function within the [`mutate` function](https://r4ds.hadley.nz/data-transform.html#sec-mutate) 


```{r}
df3 <- df3 |> mutate(year = year(datetime))

```

## Question 6


In the analysis below, you will need a column that is the month.  Apply the function [`month()`](https://r4ds.hadley.nz/datetimes.html#date-time-components) from the lubridate package to the `datetime` column and save the updated data frame. Hint: you will use the [`mutate` function](https://r4ds.hadley.nz/data-transform.html#sec-mutate) 

```{r}
df3 <- df3 |> mutate(month = month(datetime))

```

## Question 7

To finalize your cleaned dataset, order the values in the columns by Reservoir (A-Z).  Hint:  use [`arrange`](https://r4ds.hadley.nz/data-transform.html#arrange)

```{r}
df3 <- df3 |> arrange(Reservoir)

```


# Exercise 3

Exercise 3 is focused on high level summary statistics.

## Question 8

How many different reservoirs are in the dataset and what are their names? Hint: use the [`distinct`](https://r4ds.hadley.nz/data-transform.html#distinct) function

```{r}
df3 |> distinct(Reservoir)

#Answer: There are 5 different reservoirs known as "BVR", "CCR", "FCR", "GWR", and "SHR."

```

## Question 9

Which reservoir has the highest mean surface chlorophyll-a? Hint: use `group_by()` and `summarize()` [combination](https://r4ds.hadley.nz/data-transform.html#groups)

```{r}
df3 |> group_by(Reservoir) |> summarize(meanChla = mean(Chla_ugL, na.rm = TRUE))

#Answer: Reservoir GWR has the highest mean surface chlorophyll-a of 3.2247313. 

```

## Question 10

Which reservoir has the highest observed surface chlorophyll-a?

```{r}
df3 |> group_by(Reservoir) |> summarize(maxChla = max(Chla_ugL, na.rm = TRUE))

#Answer: Reservoir FCR has the highest observed surface chlorophyll-a of 258.2027. 
```

# Exercise 4

Exercise 4 dives deeper into the data.   

## Question 11

Plot the surface chlorophyll-a concentration for each year (averaged across reservoirs) and describe how surface chlorophyll-a is changing over time.

```{r}



```

## Question 12

Plot the annual mean surface chlorophyll-a concentration for each year for each reservoir and describe how the patterns differ from the plot in Question 11.  

```{r}

```

## Question 13

Plot the mean surface chlorophyll-a concentration by month for each reservoir and describe how the month of peak chlorophyll-a is similar or different across the reservoirs.

```{r}

```

# Exercise 5

Two of the reservoirs are close in distance to each other (FCR and BVR).  We are interested in how the mean annual surface chlorophyll-a differs between the two and how that difference changes over time.  

## Question 14:

Plot the difference in the annual mean chlorophyll-a between `BVR` and `FCR`.  Hints: you will need to use `group_by`, `summarize`, [`pivot_wider`](https://r4ds.hadley.nz/data-tidy.html#widening-data), and `mutate` to get the cleaned dataset from Exercise 3 into a table that can be used generate this plot.


```{r}

```

# Exercise 6

This exercise is focused on exporting the data and a plot 

## Question 15

Save the plot from Question 14 to a PNG named `question-14.png` in your repo top-level directory. Hint: use [ggsave()](https://r4ds.hadley.nz/data-visualize.html#sec-ggsave)

```{r}

```

## Question 16

Write your cleaned data frame to your repo top-level directory with the filename of `introduction-cleaned-data.csv`.  Hint use [`write_csv()`](https://r4ds.hadley.nz/data-import.html#sec-writing-to-a-file)

```{r}

```

Remember to Knit your document as a `github_document` and comment+push your code and knitted document to GitHub.
